Initial estuarine response to inorganic nutrient inputs from a legacy mining facility to Tampa Bay, Florida

### 2022-03-09 revision2/review_response.Rmd

1. The authors didn't address my question of how the plume model was verified, other than to point to a couple of papers. I read those papers and don't fully understand them as I am not a physical oceanographer, but as far as I could tell, neither paper clearly stated how verification was made using field observations. Given that this paper is likely to be read mostly by water quality experts and not physical oceanographers, I feel strongly that the authors need to include a brief (plain language… not physical oceanographer-speak) description of modeled plume verification, especially considering that the field sampling and interpretation of results is so strongly dependent on accurate quantification of plume evolution.

2. The authors speculate that nutrients were likely a driver of the abnormally high K. brevis abundances during the study. Is there any way to determine if the abundances observed in Tampa Bay were a result of growth vs. a physical concentration mechanism? This seems like it would be a good first step for giving the reader some confidence that nutrients from Piney Point played a role. For example, if you know the initial abundance coming into Tampa Bay via the inlet, can the plume model be used to estimate transit time to whatever location/time point maximum abundances were observed, and from that, can you use known growth rates to determine if those abundances are logical based on growth alone vs physical concentration?

3. The authors argue in several locations that elevated salinities are conducive to K. brevis growth. Need references to support this.

     * __Response__: We have added citations to the discussion, second to last paragraph: "Severe *K. brevis* blooms are rarer in estuaries because high abundances are most common at higher salinities typical of coastal or oceanic waters [@Steidinger98;@Villac20]."

@@

4. The discussion is awkwardly laid out, e.g., the first paragraph of "Additional interpretation of impacts" would seemingly belong in the previous section on "Nutrient cycling".

5. In the discussion section on seagrass impacts, there is a sentence that is confusing and possibly contradictory: "The results observed in 2021 suggested water quality conditions were not supportive of seagrass growth, although changes were not observed and the conditions likely did not persist long enough to impact seagrasses."

     * __Response__: This sentence was revised as "Water quality results in 2021 suggested that conditions may have been light-limiting for seagrass growth (e.g., high chlorophyll concentrations, low secchi observations), although the conditions likely did not persist long enough to impact seagrasses."

### 2022-03-09 manu-draft.Rmd
working on reviewer two comments

Previous research for Tampa Bay has identified water quality conditions that are likely to promote seagrass growth [@Greening06;@Greening14, and references therein]. The results observed in 2021 suggested water quality conditions were not supportive of seagrass growth, although changes were not observed and the conditions likely did not persist long enough to impact seagrasses.  The long-term effects of the Piney Point discharge on the seagrass community remains uncertain. From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south (Southwest Florida Water Management District, unpublished results).  These broader trends suggest regional drivers are affecting seagrass communities [e.g., variation in precipitation, @Tomasko20], yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats. Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent *K. brevis* bloom in 2018. Although the 2021 red tide in Tampa Bay was short-lived, potential long-term effects on seagrasses remain a concern [e.g., alteration of sediment geochemistry, @Eldridge04]. Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from annual transect monitoring results for Tampa Bay.  In particular, increasing abundance in recent years of the green algae *Caulerpa* sp. has been observed at long-term transects that were previously dominated by seagrass.  These changes may be indicative of broader ecosystem shifts concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06]. Acute stressors from short-term events, such as unanticipated releases from Piney Point, create additional and often preventable challenges to managing seagrass health.

@@

Establishing causal linkages between the nutrient inputs from Piney Point and the severity of the *K. brevis* bloom observed in Tampa Bay this year is difficult in the absence of more quantitative results or mechanistic tools to support understanding.  Occurrence of this species has historically been spatially distinct, with blooms originating in subsurface water offshore on the West Florida Shelf [@Steidinger75;@Weisberg14;@Weisberg19;@Liu16] and occasionally occurring at bloom concentrations in lower and middle Tampa Bay. Although bloom concentrations in 2021 were extreme, historical blooms have been observed in Tampa Bay with notable events occurring in 1971 [@Steidinger72], 2005 [@Flaherty11], and recently in 2018 [@Skripnikov21].  Seasonal persistence in Gulf waters in southwest Florida can vary between years, with some blooms lasting as short as a few weeks, while others have been present for longer than a year [the 2018 bloom lasted sixteen months, @Skripnikov21]. Severity in estuaries is typically less than Gulf waters as *K. brevis* is limited to higher salinity.  Contributing factors in 2021, such as low rainfall preceding the bloom and varying wind patterns, created conditions that were favorable for growth of *K. brevis* in Tampa Bay. However, the results suggest a likely scenario that residual nutrients from the Piney Point release, or indirectly through nutrients made available from the growth and decomposition of other primary producers (e.g., diatoms, macroalgae) stimulated by inputs from Piney Point, were sufficiently available to allow growth of *K. brevis* to the concentrations observed in July [also see @Medina20].  Daily simulation results from the Tampa Bay Coastal Ocean Model [@Chen18;@Chen19] suggested that the plume was widespread throughout the bay and persisted for many months after the release ceased at Port Manatee.  Plume dispersal also suggested that both open-water and back-bay habitats were exposed to nutrient concentrations sufficient to stimulate phytoplankton production. Although Piney Point did not cause red tide (i.e., it originates in the Gulf of Mexico), the events of 2021 created conditions in Tampa Bay conducive for the extreme bloom concentrations observed in July.

>>>>

Previous research for Tampa Bay has identified water quality conditions that are likely to promote seagrass growth [@Greening06;@Greening14, and references therein]. Water quality results in 2021 suggested that conditions may have been light-limiting for seagrass growth (e.g., high chlorophyll concentrations, low secchi observations), although the conditions likely did not persist long enough to impact seagrasses.  The long-term effects of the Piney Point discharge on the seagrass community remains uncertain. From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south (Southwest Florida Water Management District, unpublished results).  These broader trends suggest regional drivers are affecting seagrass communities [e.g., variation in precipitation, @Tomasko20], yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats. Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent *K. brevis* bloom in 2018. Although the 2021 red tide in Tampa Bay was short-lived, potential long-term effects on seagrasses remain a concern [e.g., alteration of sediment geochemistry, @Eldridge04]. Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from annual transect monitoring results for Tampa Bay.  In particular, increasing abundance in recent years of the green algae *Caulerpa* sp. has been observed at long-term transects that were previously dominated by seagrass.  These changes may be indicative of broader ecosystem shifts concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06]. Acute stressors from short-term events, such as unanticipated releases from Piney Point, create additional and often preventable challenges to managing seagrass health.

@@

Establishing causal linkages between the nutrient inputs from Piney Point and the severity of the *K. brevis* bloom observed in Tampa Bay this year is difficult in the absence of more quantitative results or mechanistic tools to support understanding.  Occurrence of this species has historically been spatially distinct, with blooms originating in subsurface water offshore on the West Florida Shelf [@Steidinger75;@Weisberg14;@Weisberg19;@Liu16] and occasionally occurring at bloom concentrations in lower and middle Tampa Bay. Although bloom concentrations in 2021 were extreme, historical blooms have been observed in Tampa Bay with notable events occurring in 1971 [@Steidinger72], 2005 [@Flaherty11], and recently in 2018 [@Skripnikov21].  Seasonal persistence in Gulf waters in southwest Florida can vary between years, with some blooms lasting as short as a few weeks, while others have been present for longer than a year [the 2018 bloom lasted sixteen months, @Skripnikov21]. Severe *K. brevis* blooms are rarer in estuaries because high abundances are most common at higher salinities typical of coastal or oceanic waters [@Steidinger98;@Villac20].  Contributing factors in 2021, such as low rainfall preceding the bloom and varying wind patterns, created conditions that were favorable for growth of *K. brevis* in Tampa Bay. However, the results suggest a likely scenario that residual nutrients from the Piney Point release, or indirectly through nutrients made available from the growth and decomposition of other primary producers (e.g., diatoms, macroalgae) stimulated by inputs from Piney Point, were sufficiently available to allow growth of *K. brevis* to the concentrations observed in July [also see @Medina20].  Daily simulation results from the Tampa Bay Coastal Ocean Model [@Chen18;@Chen19] suggested that the plume was widespread throughout the bay and persisted for many months after the release ceased at Port Manatee.  Plume dispersal also suggested that both open-water and back-bay habitats were exposed to nutrient concentrations sufficient to stimulate phytoplankton production. Although Piney Point did not cause red tide (i.e., it originates in the Gulf of Mexico), the events of 2021 created conditions in Tampa Bay conducive for the extreme bloom concentrations observed in July.
